HMMotion: Using tracking data to predict coverage strategies in American Football

### 2025-01-04 submission.Rmd
.

##### Feature engineering
To accurately forecast the defensive scheme (man- or zone defense) for every play, we need to create various features derived from the tracking data. In particular, we conducted the following feature engineering steps: We first consider all 11 players on each side of the field and compute features related to the convex hull of the positions of the players. In particular, for defense and offense, we compute the area spanned by the convex hull of all player such as well as the largest $y$ distance (i.e.\ the width of the hull) and the largest $x$ distance (i.e. the length of the hull). In addition, we select the five most relevant players on each side of the field. For offense, we omit the offensive line and the QB, while, for defense, we omit nose tackles, defensive tackles and defensive ends, and select the five defenders that were the closest to the five attackers corresponding to a weighted euclidean distance, putting much more emphasis on the y-axis. Finally, we use their standardized $x$ and $y$ coordinates as features and order defensive and offensive players according to their $y$ coordinates. Additionally, for each of the relevant defenders, we compute distances to the football and their orientation with respect to the quarterback (`r colorize("values taken at event lineset. Rouven: Die Values haben wir doch rausgenommen aus dem finalen Modell oder? Robert: ich denke das sollten wir noch besprechen. Ich habe ohnehin mehrere Modelle gefitted...","orange")`). Finally, we extract relevant information from play-by-play data, such as quarter, down, yards to go, home and away score and the game clock.

@@

We describe the movements of the five defensive players during the phase of pre-snap motion by a hidden Markov model (see the Appendix for an in-depth description). In particular, we assume that each defender's $y$-coordinate at each time point $t$ is a realization from a particular Gaussian distribution with mean according to the $y$-coordinate of the offensive player that is currently guarded and an estimated standard deviation. HMMs are particularly useful in this setting as there is no direct knowledge of the guarding allocation. This is only partially observed based on the reaction of the defender to the offenders' movement. Naturally, the model thus treats the information on who is currently guarded as a latent state variable, and hence each observation as a realization from a mixture of Gaussian distributions (see Franks et al. 2015 for a similar approach in basketball).
Ultimately, the goal is to use the fitted model for so-called *state-decoding*, i.e. inferring information on the guarding allocation based on the observations made. For this task, HMMs not only leverage information on the $y$-coordinates of the defender and all offenders potetially guarded at each time point, but also incorporate probabilistic information based on the previous and subsequent observations. For practical details on the model fitting and state decoding, also see the Appendix.

@@

#### 3. Post-motion prediction
We re-train the pre-motion model to predict whether the defense plays a man- or zone coverage scheme, but, in this step, we incorporate results from the HMM analysis as further features. However, a caveat of the aforemententioned state probabilities is that they constitute as high-dimensional time-series data which makes them hard to use them as features. An outlet of this problem is to employ suitable summary statistics (`r colorize("Robert hier spezifizieren welche du final genommen hast, ich gehe jetzt mal von summe an switches und entropy aus; Robert: Im moment nehmen ich alle 4 (sum, average, nr changes, average ent)... Sollten wir nochmal besprechen.","orange")`). In particular, we calculate for each defender the most likely offensive player $k = 1, \ldots, 5$ to be guarded in each time point $t = 1, \ldots, T_n$. From this, we count the number of state switches and sum these numbers across defenders as a first, simple, summary statistics and further calculate across defenders $j = 1, \ldots, 5$ the mean entropy in offensive play $n = 1,\ldots, N$, i.e.:

@@

While we were clearly able to show that motion enables the defensive coverage scheme, a limitation of our approach lies in the imperfect prediction accuracy of the pre-motion model, primarily due to insufficient hyperparameter tuning and the relatively small number of plays involving motion. However, the primary focus of this project was on pre-snap motion, particularly on how to effectively translate this information into the hidden Markov model. Importantly, our pre-motion model is modular and can be seamlessly replaced by another model, such as the NFL Next Gen Stats model, which can then be integrated into our modeling extension to fully leverage the insights provided by our extention.

@@

A hidden Markov model consists of an observed time series $\{\boldsymbol{y}_t\}_{t=1}^T$ and an unobserved first-order Markov chain $\{ g_t\}_{t=1}^T$, with $g_t \in \{1,\ldots,N\}$. In this case, at every time point $t$, $\boldsymbol{y}_t$ is the y-coordinate of the defensive player and $g_t$ proxies the offensives player to be guarded, i.e. the guarding allocation. 
The Markov chain is fully described by an initial distribution $\boldsymbol{\delta}^{(1)} = \bigl( \Pr(g_1=1), \ldots, \Pr(g_1=N) \bigr)$ and a transition probability matrix (t.p.m.) $\boldsymbol{\Gamma} = (\gamma_{ij})$, with $\gamma_{ij} = \Pr(g_t = j \mid g_{t-1} = i), \ i,j = 1, \ldots, N$. The connection of both stochastic processes arises from the assumption that the distribution of the observations $\boldsymbol{y}_t$ are fully determined by the state that is currently active. More formally,
\begin{equation*}
f(\boldsymbol{y}_t \mid g_1, \ldots, g_T, \boldsymbol{y}_1, \ldots, \boldsymbol{y}_{t-1},\boldsymbol{y}_{t+1},\ldots,\boldsymbol{y}_T) = f(\boldsymbol{y}_t \mid g_t = j), \qquad j \in \{1, \ldots, N\},
\end{equation*}
which we denote by $f_j(\boldsymbol{y}_t)$ in short.
In general, $f_j$ can be any density or probability mass function depending on the type of data and a typical choice is a parametric distribution with separate parameters for each latent state. Following the approaches of Franks et al. (2015), we opt for a Gaussian distribution with a mean that is fully determined by the current $y$-coordinate of defender $j \in \{1, \dots, N\}$ and a standard deviation that is fixed accross all states but estimated from the data.

To fit the model, we use direct numerical likelihood maximization. The HMM likelihood for the motion of a specific defender in a specific play can be calculated based on the so-called *forward algorithm*. It effectively performs a summation over all possible latent state sequences in an efficient manner, rendering the computational complexity linear in the number of observations. Time series of different defenders within the same play, and of different plays are treated as independent, hence their likelihood contributions are summed to obtain the full likelihood of the training data. For practical implementation, we wrote a custom likelihood function in R, using the function `forward()` and other convenience functions from the R package `LaMa` (Koslik, 2024) to speed up computations. Furthermore, we used the R package `RTMB` to make this likelihood function compatible with automatic differentiation, making the numerical optimization process more efficient and robust. 
The parameters to be estimated are only the transition probability matrix $\boldsymbol{\Gamma}$ and the standard deviation of the Gaussian distribution. 
The initial distribution of each defender for each play is fixed and computed based on ... ROUVEN

>>>>

##### Feature engineering
To accurately forecast the defensive scheme (man- or zone defense) for every play, we need to create various features derived from the tracking data. In particular, we conducted the following feature engineering steps: We first consider all 11 players on each side of the field and compute features related to the convex hull of the positions of the players. In particular, for defense and offense, we compute the area spanned by the convex hull of all player such as well as the largest $y$ distance (i.e. the width of the hull) and the largest $x$ distance (i.e. the length of the hull). In addition, we select the five most relevant players on each side of the field. For offense, we omit the offensive line and the QB, while, for defense, we omit nose tackles, defensive tackles and defensive ends, and select the five defenders that were the closest to the five attackers corresponding to a weighted euclidean distance, putting much more emphasis on the y-axis. Finally, we use their standardized $x$ and $y$ coordinates as features and order defensive and offensive players according to their $y$ coordinates. Additionally, for each of the relevant defenders, we compute distances to the football and their orientation with respect to the quarterback (`r colorize("values taken at event lineset. Rouven: Die Values haben wir doch rausgenommen aus dem finalen Modell oder? Robert: ich denke das sollten wir noch besprechen. Ich habe ohnehin mehrere Modelle gefitted...","orange")`). Finally, we extract relevant information from play-by-play data, such as quarter, down, yards to go, home and away score and the game clock.

@@

We describe the movements of the five defensive players during the phase of pre-snap motion by a hidden Markov model (see the Appendix for an in-depth description). In particular, we assume that each defender's $y$-coordinate at each time point $t$ is a realization from a particular Gaussian distribution with mean according to the $y$-coordinate of the offensive player that is currently guarded and an estimated standard deviation. HMMs are particularly useful in this setting as there is no direct knowledge of the guarding allocation. This is only partially observed based on the reaction of the defender to the offenders' movement. Naturally, the model thus treats the information on who is currently guarded as a latent state variable, and hence each observation as a realization from a mixture of Gaussian distributions (see Franks et al. 2015 for a similar approach in basketball).
Ultimately, the goal is to use the fitted model for so-called *state-decoding*, i.e. inferring information on the guarding allocation based on the observations made. For this task, HMMs not only leverage information on the $y$-coordinates of the defender and all offenders potentially guarded at each time point, but also incorporate probabilistic information based on the previous and subsequent observations. For practical details on the model fitting and state decoding, also see the Appendix.

@@

#### 3. Post-motion prediction
We re-train the pre-motion model to predict whether the defense plays a man- or zone coverage scheme, but, in this step, we incorporate results from the HMM analysis as further features. However, a caveat of the aforemententioned state probabilities is that they constitute as high-dimensional time-series data which makes them hard to use them as features. An outlet of this problem is to deduce suitable summary statistics that can serve as features in the existing model(`r colorize("Robert hier spezifizieren welche du final genommen hast, ich gehe jetzt mal von summe an switches und entropy aus; Robert: Im moment nehmen ich alle 4 (sum, average, nr changes, average ent)... Sollten wir nochmal besprechen.","orange")`). To overcome this, we first calculate for each defender the most likely offensive player $k = 1, \ldots, 5$ to be guarded in each time point $t = 1, \ldots, T_n$ and count the number of state switches, i.e. the number of times for which a defender's most likely guarded offensive players differs between two consecutive time points $s$ and $t$, $s>t$. From this, we can calculate some simple, summary statistics: 1) the sum of state switches, 2) the average number of state switches and 3) the number of defenders during a play that switch offensive players during a play. Besides, we use the aforementioned state decodings to calculate a more elaborated statistic, in particular the mean entropy across defenders $j = 1, \ldots, 5$ in each offensive play $n = 1,\ldots, N$, i.e.:

@@

While we were clearly able to show that motion facilitates the defensive coverage scheme detection, a limitation of our approach lies in the imperfect prediction accuracy of the pre-motion model, primarily due to insufficient hyperparameter tuning and the relatively small number of plays involving motion. However, the primary focus of this project was on pre-snap motion, particularly on how to effectively translate this information into the hidden Markov model. Importantly, our pre-motion model is modular and can be seamlessly replaced by another model, such as the NFL Next Gen Stats model, which can then be integrated into our modeling extension to fully leverage the insights provided by our extension.

@@

A hidden Markov model consists of an observed time series $\{\boldsymbol{y}_t\}_{t=1}^T$ and an unobserved first-order Markov chain $\{ g_t\}_{t=1}^T$, with $g_t \in \{1,\ldots,N\}$. In this case, at every time point $t$, $\boldsymbol{y}_t$ is the y-coordinate of the defensive player and $g_t$ proxies the offensives player to be guarded, i.e. the guarding allocation. 
The Markov chain is fully described by an initial distribution $\boldsymbol{\delta}^{(1)} = \bigl( \Pr(g_1=1), \ldots, \Pr(g_1=N) \bigr)$ and a transition probability matrix (t.p.m.) $\boldsymbol{\Gamma} = (\gamma_{ij})$, with $\gamma_{ij} = \Pr(g_t = j \mid g_{t-1} = i), \ i,j = 1, \ldots, N$. The connection of both stochastic processes arises from the assumption that the distribution of the observations $\boldsymbol{y}_t$ are fully determined by the state that is currently active. More formally,
\begin{equation*}
f(\boldsymbol{y}_t \mid g_1, \ldots, g_T, \boldsymbol{y}_1, \ldots, \boldsymbol{y}_{t-1},\boldsymbol{y}_{t+1},\ldots,\boldsymbol{y}_T) = f(\boldsymbol{y}_t \mid g_t = j), \qquad j \in \{1, \ldots, N\},
\end{equation*}
which we denote by $f_j(\boldsymbol{y}_t)$ in short.
In general, $f_j$ can be any density or probability mass function depending on the type of data and a typical choice is a parametric distribution with separate parameters for each latent state. Following the approaches of Franks et al. (2015), we opt for a Gaussian distribution with a mean that is fully determined by the current $y$-coordinate of defender $j \in \{1, \dots, N\}$ and a standard deviation that is fixed across all states but estimated from the data.

To fit the model, we use direct numerical likelihood maximization. The HMM likelihood for the motion of a specific defender in a specific play can be calculated based on the so-called *forward algorithm*. It effectively performs a summation over all possible latent state sequences in an efficient manner, rendering the computational complexity linear in the number of observations. Time series of different defenders within the same play, and of different plays are treated as independent, hence their likelihood contributions are summed to obtain the full likelihood of the training data. For practical implementation, we wrote a custom likelihood function in R, using the function `forward()` and other convenience functions from the R package `LaMa` (Koslik, 2024) to speed up computations. Furthermore, we used the R package `RTMB` to make this likelihood function compatible with automatic differentiation, making the numerical optimization process more efficient and robust. 
The parameters to be estimated are only the transition probability matrix $\boldsymbol{\Gamma}$ and the standard deviation of the Gaussian distribution. 
The initial distribution of each defender for each play comes from a deterministic assignment approach based on spatial proximity of the initial y-coordinates at the beginning of the time-series, i.e. the moment the line is set. Specifically, for each defender in a predefined set of potential defenders, we calculated their Euclidean distances to the five offensive players For each offense player, ranked in descending order by proximity, we assigned the closest defender to that player. The initial distribution was then set as a categorical distribution, where each defender was assigned a probability of 1 for their corresponding attacker and 0 for all others.
